Dark Control:\\ A Unified Account of Default Mode Function\\ by Markov Decision Processes

### 2017-09-18 paper/paper.tex
roundup final sentence
hide appendix

It is an important feature of the proposed artificial intelligence perspective
on DMN biology that it is practically computable and
yields falsifiable neuroscientific hypotheses.
%
Neuroscience experiments could be designed that operationalize
the set of action, value, and state variables that govern
the behavior of intelligent RL agents.
%
At the least, we propose an alternative vocabulary to
describe, contextualize, and interpret experimental findings in neuroscience studies
on the DMN.
%
Ultimately,
DMN activity may instantiate a holistic integration
ranging from real experience over purposeful dreams to anticipated futures
for continued refinement of the organism's fate. Indeed, a major hurdle in explaining DMN activity from brain-imaging studies has been its very similar engaged across time scales: thinking about the past (e.g., autobiographical memory), thinking about hypothetical presents (e.g., dreams during sleep or daytime mind-wandering), and thinking about future scenarios (e.g., prospection or delay discounting). The MDP account of DMN function naturally integrates this a-priori diverging mental processes into a common framework. % It is this feature of our functional view that we underline in the referred sentence.

@@

\appendix

\begin{remark}
  Given a system $(\mathcal S, \mathcal A, r, p, \mu)$ satisfying all the axioms for an MDP
  except the Markov property, we can always transform it into an MDP by considering a compounded
  version of the states
  $$S_t \leftarrow (S_0,A_0,R_0,S_1,A_1,S_2\ldots,S_t,A_{t-1},R_{t-1},S_{t})$$ made of the system's
  history up to an including time $t$.
\end{remark}

\section{More on RL in the brain}
\subsection{Does hippocampal replay equal inverse reinforcement-learning?}
%\begin{mdframed}
  Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
  an MDP (called a \textit{teacher's demonstration}), can we figure out what is the
  (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
  \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
  (e.g., linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
  For instance, given traces of motor actions from an adult to grab a cup on a table,
  can an observing child figure out what ``langragian'' functional is being minimized by the former?
  How? Can they reproduce this optimal behavior?
%\end{mdframed}
Such questions are of course pertinent to our decision-making theory for the DMN.
In the general artificial case, the problem has been extensively studied and partially solved by
\citep{abbeel2004}. They are also been rigorously studied in general optimal control literature under the name
``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...

IRL is is suited for problems in which it's hard to define what the reward function could be
(e.g.,  car-driving, drone maneuvering, etc.) ...

\section{Free-energy principles!}
The so-called free-energy principle in its present form (including notions like ``generative density'', ``recognition density'', etc.) can be traced back to works of Dayan \& Hinton \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial directional deep belief-net (DBN).
In this subsection, we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in building their Helmholtz machine. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For instance, the Friston theory axiomatizes that the brain uses a (problematic, as it implicitly assumes that posterior of each hidden unit is factorial) wake-sleep algorithm to train the underlying Helmholtz machine, etc.

% \begin{mdframed}
%   Though our presentation of free-energy principles is thermodynamic in flavor, a completely equivalent
%   information-theoretic formulation is of course possible, simply by exploiting the information-theoretic interpretation of entropy.
% \end{mdframed}

\subsection{Helmholtz free-energy and the generative model}
\begin{table}[H]
  \begin{tabular}{p{2cm}|p{11cm}}
         \hline
         \textbf{symbol}    & \textbf{description}  \\ \hline
         $\langle X\rangle_p$ & Expectation (a.k.a average, a.k.a mean value) of the
         random quantity $X$ w.r.t to the probability density $p$, formally defined by $\langle E\rangle_p := \sum_{z}p(z)X(z)$.\\ \hline
         $\mathcal H(p)$ & Information-theoretic entropy of a probability density $p$, formally defined by $\mathcal H(p) := -\sum_{z}p(z)\log(p(z))$,
          with the usual convention $0 \log(0) := 0$.\\ \hline
         $D_{KL}(q||p)$ & The Kullback-Leibler divergence between the probability densities $q$ and $p$ respectively, formally defined by $D_{KL}(q||p) := \sum_{z}q(z)\log(q(z)/p(z))$.\\ \hline
             $\x$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's internal state $b$ and sensory inputs $s$, i.e., $\x = (s, b).$ \\ \hline
             $\z$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
             $p_G(.|\x)$ & Generative density for ...\\ \hline
         $p_R(.|\x)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
         $F_G(\x)$ & Helmholtz free-energy for a model $p_G$ of generating the observation $\x$. This measures the surprise incured upon observing $\x$ generated by the model $G$.\\ \hline
         $F^R_G(\x)$ & Variational Helmholtz free-energy from model $G$
          to $R$.  Note that $F^G_G = F_G$.\\ \hline
  \end{tabular}
  \caption{Table of notations.}
\end{table}
Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
\begin{eqnarray}
  \begin{split}
    \text{generative surprise } &= -\log(p_G(\x)) = -\log(p_G(\x)) \times 1 = -\log(p_G(\x))\sum_{\z}p_G(\z |\x)\\
    &= -\sum_{\z}p_G(\z, \x)\log(p_G(\x))
    =-\sum_{\z}p_G(\z |\x)\log(p_G(\z, \x)/p_G(\z|\x))\\
    &= \sum_{\z}p_G(\z |\x)\log(p_G(\z|\x))-\sum_{\z}p_G(\z |\x)\log  (p_G(\z, \x))\\
    &= -\langle \log  (p_G(., \x)) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x))\\
    &= \langle E_G(., \x) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x))
  \end{split}
  \label{eq:helm}
\end{eqnarray}
where $E_G(\z, \x)$ is the energy at \textit{macrostate} $\z$ of a fictive thermodynamic system defined by setting
\begin{equation}
  E_G(\z, \x) := -\log(p_G(\z, \x)),
  \label{eq:gibbs}
\end{equation}
%% In the above, the variable $\z$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
The last quantity in \eqref{eq:helm} is nothing but \textit{Helmholtz free-energy} (at unit temperature!), defined formally by
\begin{equation}
  F_G(\x) := \langle E_G(., \x) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x)).
\end{equation}
%% , modelled as the physical system for which $p_G(\z|\x)$ dictates the occupation probabilities of macrostates $\z$
Thus,
\begin{fact}
  Generative surprise and generative Helmholtz free-energy are different views on exactly the same object.
\end{fact}

The goal of the brain is then to optimize over the generative model $G$: to iteratively or analytically modify the generative density $p_G(.|\x)$, so as to minimize surprise. It turns out that a direct attempt to attack this optimization problem by gradient descent on the free-energy $F_G(\x)$ is furtile: the parameter update steps are not ``very clean'', and require rather cumbersome and heavy computations. A workaround is then to introduce a second density $p_R(.|\x)$ called a \textit{recognition} density to work in tandem with the generative density $p_G(.|\x)$, as a trick for doing approximate inference. The former dreams / fantacizes whilst the latter tries to generate sensations which match these dreams! This primal-dual idea, first proposed in Hinton et al. 1995, is at the heart of the general free-energy principle that we will introduce shortly.

\subsection{Variational Helmholtz free-energy and the bottom-up recognition sub-model}
In this subsection, we will present an insightful upper bound for the generative surprise (i.e., generate Helmholtz free-energy), called the \textit{variational} (Helmholtz) free-energy. As an avant-gout of what is to come shorty, let's just note that the well-known \textit{free-energy principle} is simply a workaround whereby the minimization surprise (intractable) is replaced with the  minimization a carefully chosen upper bound thereof.

Invoking \eqref{eq:gibbs} and applying Bayes rule, we get the Gibbs
distribution
\begin{equation}
  p_G(\z|\x) = \frac{p_G(\z|\x)}{p_G(\x)} = \frac{\exp(-E_G(\z, \x))}{Z_G(\x)} = \frac{\exp(-E_G(\z, \x))}{Z_G(\x)},
\end{equation}
where $Z_G(\x) := \log(p_G(\x)) = \sum_{\z'}\exp(-E_G(\z',\x))$, the normalizing \textit{partition function} for the model \ref{eq:gibbs}.
Whence, for any macrostate $\z$, we have $p_G(\x) = Z_G(\x) = \exp(-E_G(\z, \x)) / p_G(\z|\x)$, and so it holds that
\begin{equation}
F_G(\x)\overset{\eqref{eq:helm}}{=} -\log(p_G(\x)) = -\log(Z_G(\x)) = E_G(\z, \x) + \log(p_G(\z|\x)).
\end{equation}
Now, in the above equation, the LHS only depends on the generative model $G$ and the data point $\x$:
it doesn't depend on the hidden variable $\z$, etc. So, taking expectations w.r.t an arbitrary
density\footnote{conditioning in $p_R(.|\x)$ is because this density is selected from a world in which the sensory inputs and internal brain
  state vector $\x$ is assumed already observed.} $p_R(.|\x)$ yields
\begin{equation}
  \begin{split}
    F_G(\x) &= -\log(Z_G(\x)) = \langle E_G(., \x)\rangle_{p_R(.|\x)} + \sum_{\z}p_R(\z|\x)\log(p_G(\z|\x))\\
    &= \langle E_G(., \x)\rangle_{p_R(.|\x)} - \mathcal H(p_R(.|\x)) - \sum_{\z}p_R(\z|\x)\log(p_R(\z|\x)/p_G(\z|\x))\\
    &= F^R_G(\x) - D_{KL}(p_R(.|\x) || p_G(.|\x)),
  \end{split}
  \label{eq:fe}
\end{equation}
where $F^R_G(\x)$ is the \textit{variational} Helmholtz free-energy from $R$ to $G$ defined by
\begin{equation}
  F^R_G(\x) := \langle E_G(., \x)\rangle_{p_R(.|\x)} - \mathcal H(p_R(.|\x))
\end{equation}
and $D_{KL}(p_R(.|\x) || p_G(.|\x))$ is the Kullback-Leibler divergence between the $p_R(.|\x)$ and the generative density $p_G(.|\x)$. Note that $F^G_G = F_G$.

\subsection{A general free-energy principle}
We can resume the situation as follows\footnote{Where we have used the fact that KL divergence is always nonnegative.}:
\begin{mdframed}
\begin{equation}
  \begin{split}
    \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
    &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{D_{KL}(p_R(.|\x) || p_G(.|\x))}_{\text{complexity}} \\
    &\le F^R_G(\x),
    \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z
    \end{split}
\end{equation}
\end{mdframed}
%% In fact, we have the following theorem
%% \begin{theorem}
%%   It holds that
%%   $$
%%   \text{Minimal generative surprise = minimal (Helmholtz) variational free-energy},$$
%%   i.e
%%   \begin{equation}
%%     \min_{G}F_G(\x) = \min_{G,R}F^R_G(\x)
%%     \end{equation}
%% \end{theorem}

\subsection{Helmholtz machines and the wake-sleep algorithm}
% \begin{mdframed}
  \textbf{Assumption:} In both generative and recognition components of the network, there is conditional independence of  neurons in the same layer, given the data (i.e., input from lower more primitive layers). Precisely
$$
p_G(\z^{(l)}|\x) = \Pi_{k=1}^{h_l}p_G(\z_k^{(l)} | \x),\hspace{1cm}
p_R(\z^{(l)}|\x) = \Pi_{k=1}^{h_l}p_R(\z_k^{(l)} |\x)
$$
% \end{mdframed}

\subsection{Friston's active-inference and agency}
\label{sec:friston}
This is nothing but an application of the Dayan's wake-sleep algorithm for training a Helmholtz machine model of the brain...

The following critics can be made:
\begin{itemize}
  \item
    As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence assumption which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
  \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
  \item Also, note that the wake-sleep algorithm doesn't do backprop! This is due to
    technical difficulty in getting derivatives of loss function w.r.t
    recognition weights $\W^R$).
  \item This difficulty was removed in the 2010s by \citep{kingma2013auto}, an other groups, via a ``reparametrization trick''.
    \end{itemize}

\subsection{Minimizing free-energy via backprop: variational auto-encoders}

% \subsection{Free-energy minimization via variational auto-encoders}
\label{sec:vae}
Here, we present a way to alleviate some conceptual and computational issues with the
free-energy framework presented thus far, by using the recent \textit{variational auto-encoder} (VAE) theory \citep{kingma2013auto}. Define the data-dependent
auxiliary random function
\begin{eqnarray}
  f_{G,R}(., \x) :\z \mapsto \log(p_G(\z,\x)) - \log(p_R(\z|\x)).
\end{eqnarray}
Then we can rewrite the
variational free-energy as
\begin{eqnarray*}
  \begin{split}
    F_G^R(\x) &:= \langle E_G(., \x) \rangle_{p_R(. |\x)} - \mathcal H(p_R(. |\x)) = \langle E_G(., \x) + \log(p_R(.|\x))\rangle_{p_R(.|\x)}\\
    &=\langle -\log(p_G(.,\x)) + \log(p_R(.|\x))\rangle_{p_R(.|\x)}\\
    &= -\langle f_{G,R}\rangle_{p_R(.|\x)} \approx -\frac{1}{M}\sum_{m=1}^Mf_{G,R}(\z^{(m)}), \text{ with }\z^{(1)},\ldots,\z^{(M)} \sim p_R(.|\x), \text{ and }M \rightarrow \infty.
    \end{split}
\end{eqnarray*}

\begin{mdframed}
  \textbf{Problem:} How do we sample from the recognition density $p_R(.|\x)$ in such a way that the sampling process is differentiable w.r.t the weights of the recognition network $\W^R$ ?
\end{mdframed}

% \begin{figure}[!tb]
%   \centering
%   \includegraphics[width=.8\linewidth]{vae.pdf}
%   \caption{Variational autoencoders...}
%   \label{fig:vae}
% \end{figure}

\paragraph{Solution: the reparametrization trick.}
\begin{itemize}
\item Choose $\epsilon \sim p_{\text{noise}}$ (noise distribution, independent of $\W^R$!)
\item Set $\z = g(\W^{R}, \x, \epsilon)$, where $g$ is an appropriate class $\mathcal C^1$ function
  \begin{itemize}
  \item results in a sample $\z \sim p_R(.|\x)$, from the correct
    posterior
    \end{itemize}
\end{itemize}
The mapping $g$ should be taught of as a ``blurring'' function which produces noisy versions $\z$,
called \textit{sensations}, of the true world state $\x$.
The result is a scheme for training DBNs via good-old backprop!
Refer to Fig. vae.pdf.
% \ref{fig:vae}.
Some examples of the reparametrization trick for a number of
choices of the posterior distribution are given in Tab. \ref{tab:rptrick}.

\begin{table}[H]
  \begin{tabular}{p{1.7cm}|p{1.5cm}|p{1.9cm}|p{2.1cm}|p{5cm}}
         \hline
         Posterior & $p_R(.|\x)$ & noise & $g(\W^R,\x,\epsilon)$ & Also \\ \hline
         Normal & $\mathcal N(\mu,\sigma)$ & $\epsilon \sim \mathcal N(0, 1)$ & $\mu + \sigma \odot \epsilon$ & Location-scale family: Laplace, Elliptical,
         Student’s t, Logistic, Uniform, Triangular, \ldots \\ \hline
         Exponential & $\mathcal \exp(\lambda)$ & $\epsilon \sim \mathcal U([0, 1))$ & $-\log(1-\epsilon)/\lambda$ & Invertible CDF: Cauchy, Logistic, Rayleigh, Pareta, Weibull, Reciprocal, Gompert, Gumbel, Erlan, ... \\ \hline
         Other & $\log\mathcal N(\mu,\sigma)$ & $\epsilon \sim \mathcal N(0, 1)$ & $\exp(\mu + \sigma \odot \epsilon)$ & Gamma, Dirichlet, Beta, Chi-squared, F, ... \\ \hline
  \end{tabular}
  \caption{Reparametrization trick \citep{kingma2013auto} for a variety of models.}
  \label{tab:rptrick}
\end{table}

For fixed $p_G(.|\x)$, the optimal choice for $p_R(.|\x)$ is given analytically by
\begin{equation}
  p_R(\z|\x) \propto p_G(\z|\x)\exp(-\Delta E_{G \rightarrow R}(\z, \x)).
\end{equation}

>>>>

It is an important feature of the proposed artificial intelligence perspective
on DMN biology that it is practically computable and
yields falsifiable neuroscientific hypotheses.
%
Neuroscience experiments could be designed that operationalize
the set of action, value, and state variables that govern
the behavior of intelligent RL agents.
%
At the least, we propose an alternative vocabulary to
describe, contextualize, and interpret experimental findings in neuroscience studies
on the DMN.
%
Ultimately,
DMN activity may instantiate a holistic integration
ranging from real experience over purposeful dreams to anticipated futures
for continued refinement of the organism's fate. Indeed, a major hurdle in explaining DMN activity from brain-imaging studies has been its very similar engaged across time scales: thinking about the past (e.g., autobiographical memory), thinking about hypothetical presents (e.g., dreams during sleep or daytime mind-wandering), and thinking about future scenarios (e.g., prospection or delay discounting). The MDP account of DMN function offers a natural integration of these a-priori diverging mental processes into a common framework. % It is this feature of our functional view that we underline in the referred sentence.

@@

% \appendix
% 
% \begin{remark}
%   Given a system $(\mathcal S, \mathcal A, r, p, \mu)$ satisfying all the axioms for an MDP
%   except the Markov property, we can always transform it into an MDP by considering a compounded
%   version of the states
%   $$S_t \leftarrow (S_0,A_0,R_0,S_1,A_1,S_2\ldots,S_t,A_{t-1},R_{t-1},S_{t})$$ made of the system's
%   history up to an including time $t$.
% \end{remark}
% 
% \section{More on RL in the brain}
% \subsection{Does hippocampal replay equal inverse reinforcement-learning?}
% %\begin{mdframed}
%   Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
%   an MDP (called a \textit{teacher's demonstration}), can we figure out what is the
%   (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
%   \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
%   (e.g., linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
%   For instance, given traces of motor actions from an adult to grab a cup on a table,
%   can an observing child figure out what ``langragian'' functional is being minimized by the former?
%   How? Can they reproduce this optimal behavior?
% %\end{mdframed}
% Such questions are of course pertinent to our decision-making theory for the DMN.
% In the general artificial case, the problem has been extensively studied and partially solved by
% \citep{abbeel2004}. They are also been rigorously studied in general optimal control literature under the name
% ``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...
% 
% IRL is is suited for problems in which it's hard to define what the reward function could be
% (e.g.,  car-driving, drone maneuvering, etc.) ...
% 
% \section{Free-energy principles!}
% The so-called free-energy principle in its present form (including notions like ``generative density'', ``recognition density'', etc.) can be traced back to works of Dayan \& Hinton \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial directional deep belief-net (DBN).
% In this subsection, we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in building their Helmholtz machine. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For instance, the Friston theory axiomatizes that the brain uses a (problematic, as it implicitly assumes that posterior of each hidden unit is factorial) wake-sleep algorithm to train the underlying Helmholtz machine, etc.
% 
% % \begin{mdframed}
% %   Though our presentation of free-energy principles is thermodynamic in flavor, a completely equivalent
% %   information-theoretic formulation is of course possible, simply by exploiting the information-theoretic interpretation of entropy.
% % \end{mdframed}
% 
% \subsection{Helmholtz free-energy and the generative model}
% \begin{table}[H]
%   \begin{tabular}{p{2cm}|p{11cm}}
%          \hline
%          \textbf{symbol}    & \textbf{description}  \\ \hline
%          $\langle X\rangle_p$ & Expectation (a.k.a average, a.k.a mean value) of the
%          random quantity $X$ w.r.t to the probability density $p$, formally defined by $\langle E\rangle_p := \sum_{z}p(z)X(z)$.\\ \hline
%          $\mathcal H(p)$ & Information-theoretic entropy of a probability density $p$, formally defined by $\mathcal H(p) := -\sum_{z}p(z)\log(p(z))$,
%           with the usual convention $0 \log(0) := 0$.\\ \hline
%          $D_{KL}(q||p)$ & The Kullback-Leibler divergence between the probability densities $q$ and $p$ respectively, formally defined by $D_{KL}(q||p) := \sum_{z}q(z)\log(q(z)/p(z))$.\\ \hline
%              $\x$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's internal state $b$ and sensory inputs $s$, i.e., $\x = (s, b).$ \\ \hline
%              $\z$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
%              $p_G(.|\x)$ & Generative density for ...\\ \hline
%          $p_R(.|\x)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
%          $F_G(\x)$ & Helmholtz free-energy for a model $p_G$ of generating the observation $\x$. This measures the surprise incured upon observing $\x$ generated by the model $G$.\\ \hline
%          $F^R_G(\x)$ & Variational Helmholtz free-energy from model $G$
%           to $R$.  Note that $F^G_G = F_G$.\\ \hline
%   \end{tabular}
%   \caption{Table of notations.}
% \end{table}
% Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
% \begin{eqnarray}
%   \begin{split}
%     \text{generative surprise } &= -\log(p_G(\x)) = -\log(p_G(\x)) \times 1 = -\log(p_G(\x))\sum_{\z}p_G(\z |\x)\\
%     &= -\sum_{\z}p_G(\z, \x)\log(p_G(\x))
%     =-\sum_{\z}p_G(\z |\x)\log(p_G(\z, \x)/p_G(\z|\x))\\
%     &= \sum_{\z}p_G(\z |\x)\log(p_G(\z|\x))-\sum_{\z}p_G(\z |\x)\log  (p_G(\z, \x))\\
%     &= -\langle \log  (p_G(., \x)) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x))\\
%     &= \langle E_G(., \x) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x))
%   \end{split}
%   \label{eq:helm}
% \end{eqnarray}
% where $E_G(\z, \x)$ is the energy at \textit{macrostate} $\z$ of a fictive thermodynamic system defined by setting
% \begin{equation}
%   E_G(\z, \x) := -\log(p_G(\z, \x)),
%   \label{eq:gibbs}
% \end{equation}
% %% In the above, the variable $\z$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
% The last quantity in \eqref{eq:helm} is nothing but \textit{Helmholtz free-energy} (at unit temperature!), defined formally by
% \begin{equation}
%   F_G(\x) := \langle E_G(., \x) \rangle_{p_G(. |\x)} - \mathcal H(p_G(. |\x)).
% \end{equation}
% %% , modelled as the physical system for which $p_G(\z|\x)$ dictates the occupation probabilities of macrostates $\z$
% Thus,
% \begin{fact}
%   Generative surprise and generative Helmholtz free-energy are different views on exactly the same object.
% \end{fact}
% 
% The goal of the brain is then to optimize over the generative model $G$: to iteratively or analytically modify the generative density $p_G(.|\x)$, so as to minimize surprise. It turns out that a direct attempt to attack this optimization problem by gradient descent on the free-energy $F_G(\x)$ is furtile: the parameter update steps are not ``very clean'', and require rather cumbersome and heavy computations. A workaround is then to introduce a second density $p_R(.|\x)$ called a \textit{recognition} density to work in tandem with the generative density $p_G(.|\x)$, as a trick for doing approximate inference. The former dreams / fantacizes whilst the latter tries to generate sensations which match these dreams! This primal-dual idea, first proposed in Hinton et al. 1995, is at the heart of the general free-energy principle that we will introduce shortly.
% 
% \subsection{Variational Helmholtz free-energy and the bottom-up recognition sub-model}
% In this subsection, we will present an insightful upper bound for the generative surprise (i.e., generate Helmholtz free-energy), called the \textit{variational} (Helmholtz) free-energy. As an avant-gout of what is to come shorty, let's just note that the well-known \textit{free-energy principle} is simply a workaround whereby the minimization surprise (intractable) is replaced with the  minimization a carefully chosen upper bound thereof.
% 
% Invoking \eqref{eq:gibbs} and applying Bayes rule, we get the Gibbs
% distribution
% \begin{equation}
%   p_G(\z|\x) = \frac{p_G(\z|\x)}{p_G(\x)} = \frac{\exp(-E_G(\z, \x))}{Z_G(\x)} = \frac{\exp(-E_G(\z, \x))}{Z_G(\x)},
% \end{equation}
% where $Z_G(\x) := \log(p_G(\x)) = \sum_{\z'}\exp(-E_G(\z',\x))$, the normalizing \textit{partition function} for the model \ref{eq:gibbs}.
% Whence, for any macrostate $\z$, we have $p_G(\x) = Z_G(\x) = \exp(-E_G(\z, \x)) / p_G(\z|\x)$, and so it holds that
% \begin{equation}
% F_G(\x)\overset{\eqref{eq:helm}}{=} -\log(p_G(\x)) = -\log(Z_G(\x)) = E_G(\z, \x) + \log(p_G(\z|\x)).
% \end{equation}
% Now, in the above equation, the LHS only depends on the generative model $G$ and the data point $\x$:
% it doesn't depend on the hidden variable $\z$, etc. So, taking expectations w.r.t an arbitrary
% density\footnote{conditioning in $p_R(.|\x)$ is because this density is selected from a world in which the sensory inputs and internal brain
%   state vector $\x$ is assumed already observed.} $p_R(.|\x)$ yields
% \begin{equation}
%   \begin{split}
%     F_G(\x) &= -\log(Z_G(\x)) = \langle E_G(., \x)\rangle_{p_R(.|\x)} + \sum_{\z}p_R(\z|\x)\log(p_G(\z|\x))\\
%     &= \langle E_G(., \x)\rangle_{p_R(.|\x)} - \mathcal H(p_R(.|\x)) - \sum_{\z}p_R(\z|\x)\log(p_R(\z|\x)/p_G(\z|\x))\\
%     &= F^R_G(\x) - D_{KL}(p_R(.|\x) || p_G(.|\x)),
%   \end{split}
%   \label{eq:fe}
% \end{equation}
% where $F^R_G(\x)$ is the \textit{variational} Helmholtz free-energy from $R$ to $G$ defined by
% \begin{equation}
%   F^R_G(\x) := \langle E_G(., \x)\rangle_{p_R(.|\x)} - \mathcal H(p_R(.|\x))
% \end{equation}
% and $D_{KL}(p_R(.|\x) || p_G(.|\x))$ is the Kullback-Leibler divergence between the $p_R(.|\x)$ and the generative density $p_G(.|\x)$. Note that $F^G_G = F_G$.
% 
% \subsection{A general free-energy principle}
% We can resume the situation as follows\footnote{Where we have used the fact that KL divergence is always nonnegative.}:
% \begin{mdframed}
% \begin{equation}
%   \begin{split}
%     \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
%     &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{D_{KL}(p_R(.|\x) || p_G(.|\x))}_{\text{complexity}} \\
%     &\le F^R_G(\x),
%     \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z
%     \end{split}
% \end{equation}
% \end{mdframed}
% %% In fact, we have the following theorem
% %% \begin{theorem}
% %%   It holds that
% %%   $$
% %%   \text{Minimal generative surprise = minimal (Helmholtz) variational free-energy},$$
% %%   i.e
% %%   \begin{equation}
% %%     \min_{G}F_G(\x) = \min_{G,R}F^R_G(\x)
% %%     \end{equation}
% %% \end{theorem}
% 
% \subsection{Helmholtz machines and the wake-sleep algorithm}
% % \begin{mdframed}
%   \textbf{Assumption:} In both generative and recognition components of the network, there is conditional independence of  neurons in the same layer, given the data (i.e., input from lower more primitive layers). Precisely
% $$
% p_G(\z^{(l)}|\x) = \Pi_{k=1}^{h_l}p_G(\z_k^{(l)} | \x),\hspace{1cm}
% p_R(\z^{(l)}|\x) = \Pi_{k=1}^{h_l}p_R(\z_k^{(l)} |\x)
% $$
% % \end{mdframed}
% 
% \subsection{Friston's active-inference and agency}
% \label{sec:friston}
% This is nothing but an application of the Dayan's wake-sleep algorithm for training a Helmholtz machine model of the brain...
% 
% The following critics can be made:
% \begin{itemize}
%   \item
%     As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence assumption which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
%   \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
%   \item Also, note that the wake-sleep algorithm doesn't do backprop! This is due to
%     technical difficulty in getting derivatives of loss function w.r.t
%     recognition weights $\W^R$).
%   \item This difficulty was removed in the 2010s by \citep{kingma2013auto}, an other groups, via a ``reparametrization trick''.
%     \end{itemize}
% 
% 
% \subsection{Minimizing free-energy via backprop: variational auto-encoders}
% 
% % \subsection{Free-energy minimization via variational auto-encoders}
% \label{sec:vae}
% Here, we present a way to alleviate some conceptual and computational issues with the
% free-energy framework presented thus far, by using the recent \textit{variational auto-encoder} (VAE) theory \citep{kingma2013auto}. Define the data-dependent
% auxiliary random function
% \begin{eqnarray}
%   f_{G,R}(., \x) :\z \mapsto \log(p_G(\z,\x)) - \log(p_R(\z|\x)).
% \end{eqnarray}
% Then we can rewrite the
% variational free-energy as
% \begin{eqnarray*}
%   \begin{split}
%     F_G^R(\x) &:= \langle E_G(., \x) \rangle_{p_R(. |\x)} - \mathcal H(p_R(. |\x)) = \langle E_G(., \x) + \log(p_R(.|\x))\rangle_{p_R(.|\x)}\\
%     &=\langle -\log(p_G(.,\x)) + \log(p_R(.|\x))\rangle_{p_R(.|\x)}\\
%     &= -\langle f_{G,R}\rangle_{p_R(.|\x)} \approx -\frac{1}{M}\sum_{m=1}^Mf_{G,R}(\z^{(m)}), \text{ with }\z^{(1)},\ldots,\z^{(M)} \sim p_R(.|\x), \text{ and }M \rightarrow \infty.
%     \end{split}
% \end{eqnarray*}
% 
% \begin{mdframed}
%   \textbf{Problem:} How do we sample from the recognition density $p_R(.|\x)$ in such a way that the sampling process is differentiable w.r.t the weights of the recognition network $\W^R$ ?
% \end{mdframed}
% 
% % \begin{figure}[!tb]
% %   \centering
% %   \includegraphics[width=.8\linewidth]{vae.pdf}
% %   \caption{Variational autoencoders...}
% %   \label{fig:vae}
% % \end{figure}
% 
% \paragraph{Solution: the reparametrization trick.}
% \begin{itemize}
% \item Choose $\epsilon \sim p_{\text{noise}}$ (noise distribution, independent of $\W^R$!)
% \item Set $\z = g(\W^{R}, \x, \epsilon)$, where $g$ is an appropriate class $\mathcal C^1$ function
%   \begin{itemize}
%   \item results in a sample $\z \sim p_R(.|\x)$, from the correct
%     posterior
%     \end{itemize}
% \end{itemize}
% The mapping $g$ should be taught of as a ``blurring'' function which produces noisy versions $\z$,
% called \textit{sensations}, of the true world state $\x$.
% The result is a scheme for training DBNs via good-old backprop!
% Refer to Fig. vae.pdf.
% % \ref{fig:vae}.
% Some examples of the reparametrization trick for a number of
% choices of the posterior distribution are given in Tab. \ref{tab:rptrick}.
% 
% \begin{table}[H]
%   \begin{tabular}{p{1.7cm}|p{1.5cm}|p{1.9cm}|p{2.1cm}|p{5cm}}
%          \hline
%          Posterior & $p_R(.|\x)$ & noise & $g(\W^R,\x,\epsilon)$ & Also \\ \hline
%          Normal & $\mathcal N(\mu,\sigma)$ & $\epsilon \sim \mathcal N(0, 1)$ & $\mu + \sigma \odot \epsilon$ & Location-scale family: Laplace, Elliptical,
%          Student’s t, Logistic, Uniform, Triangular, \ldots \\ \hline
%          Exponential & $\mathcal \exp(\lambda)$ & $\epsilon \sim \mathcal U([0, 1))$ & $-\log(1-\epsilon)/\lambda$ & Invertible CDF: Cauchy, Logistic, Rayleigh, Pareta, Weibull, Reciprocal, Gompert, Gumbel, Erlan, ... \\ \hline
%          Other & $\log\mathcal N(\mu,\sigma)$ & $\epsilon \sim \mathcal N(0, 1)$ & $\exp(\mu + \sigma \odot \epsilon)$ & Gamma, Dirichlet, Beta, Chi-squared, F, ... \\ \hline
%   \end{tabular}
%   \caption{Reparametrization trick \citep{kingma2013auto} for a variety of models.}
%   \label{tab:rptrick}
% \end{table}
% 
% For fixed $p_G(.|\x)$, the optimal choice for $p_R(.|\x)$ is given analytically by
% \begin{equation}
%   p_R(\z|\x) \propto p_G(\z|\x)\exp(-\Delta E_{G \rightarrow R}(\z, \x)).
